GPU Acceleration of the Material Point Method

### 2019-02-13 pres/pres.tex
last apge

\begin{frame}
Sorting can dramatically increase workload performance of subsequent steps for neighboring queries:
\begin{enumerate}
  \item Deep sorted accesses are now \textbf{coalesced}. \label{it:sort_1}
  \item \textbf{Data reuse} due to L2-Cache and/or shared memory. \label{it:sort_2}
\end{enumerate}
\textbf{Double buffer particles} to use last sorted state as input for new sorting to profit from item 1 and 2!
\begin{table}
  \footnotesize
  \begin{tabular}{ | l | r | r | r | r| r |r|}    \hline
    Ordering            &  $\Delta t_c (\mu s)$ & Speedup &VRAM              & SM      & L2     & L2-Hit \\\hline
    Random              &  1,516                & -       &25.0\%            & 3.4\%   & 9.1\%  & 10.8\% \\\hline
    Deep sorted           &    218                & 6.95x   &\textbf{75.3\%}   & 24.4\%  & \textbf{35.0\%} & \textbf{37.8\%} \\
    \hline
 \end{tabular}
\caption{\footnotesize Order dependency of binning of $1024\times1024$ randomly positioned particles in a $128\times128\times128$ grid.}\label{table:bin}
\end{table}
\end{frame}
\subsection{The MPM Specific Transfers}
\begin{frame}
\frametitle{The MPM Specific Transfers}
All MPM operations belong to one of those parallelization schemes:
\begin{itemize}
 \item 1 thread : 1 particle: $\square_p = \square_p \circ \square_p \circ ... \circ \square_p$.
 \item 1 thread : 1 node: $\square_p = \square_p \circ \square_p \circ ... \circ \square_p$.
 \item G2P-transfer: $\square_p = \sum_i \square_i \circ \square_{ip}.$
 \item P2G-transfer: $\square_i = \sum_p \square_p \circ \square_{ip}.$
\end{itemize}
\vfill
MPM-Transfers are executed \textbf{multiple times per physical frame} with varying numbers of variables and mathematical operations.\\\vspace{5}
$\Rightarrow$ Preprocessing steps only need to be done \textbf{once per physical frame}. Sorting already introduced as one of these.
\end{frame}
\begin{frame}
\begin{minipage}{\textwidth}
  G2P-Transfer: $\square_p = \sum_i \square_i \circ \square_{ip}.$\\
  \includesvg[svgpath =../images/,width=\textwidth]{block_transfer_inv_par}
  Pull-Operations read!
\end{minipage}
\end{frame}
\begin{frame}
Similar to filter or stencil operations on the GPU:
\begin{itemize}
  \item 1 node : 1 thread, split grid into blocks corr. to thread groups.
  \item Interpolation function however dependent on particle position.
  \item Needs to be rerun for every particle in the cell.
\end{itemize}
\vfill
Typical setup of transfers:
\begin{enumerate}
  \item Initialize shared memory (pull: with nodes from global memory).
  \item Perform transfers.
  \item Write back to global memory (push: global atomics since writes on halo).
\end{enumerate}
\end{frame}
\begin{frame}
\begin{minipage}{\textwidth}
  P2G-transfer: $\square_i = \sum_p \square_p \circ \square_{ip}.$\\
  \includesvg[svgpath =../images/,width=\textwidth]{block_transfer_par}
  Pull-Operations read, Push-operations write!\\
  $\Rightarrow$ Race conditions.
\end{minipage}
\end{frame}
\begin{frame}
  Typically simulation domain(grid) much bigger than simulation model. Filtering inactive blocks as a preprocess improves performance.
\begin{itemize}
  \item Block is active if any cell counter is active.
  \item Cell counter is active if it has at least one particle.
\end{itemize}
\vfill
\begin{minipage}{0.9\textwidth}
  \footnotesize
  \includesvg[svgpath =../images/,width=\textwidth]{indirect_dispatch}
\end{minipage}
\end{frame}
\begin{frame}
\begin{itemize}
  \item Block and its halo are always target of shared memory operations $\Rightarrow$ P2G-Pull low occupancy.
  \item Batching multiple particles can increase performance due to hiding synchronization, unroll!
  \item Transfers respect shared memory bank conflicts fully.
  \item Warp divergence for varying cell counts.
\end{itemize}

@@

Largely same decision making:\\
\begin{table}
  \begin{tabular}{|l|c|c|}
    \hline
           & Me              & \cite{Gao:2018} \\\hline
    Sort   & Count/Histogram & Count/Histogram \\
	   & for each var.   & sel. variables  \\\hline
    Filtering domain& Filter-op.& Sparse Grid structure \\\hline
    Transfers        & Shared mem. only & Warp-shuffle op.\\\hline
  \end{tabular}
\end{table}
Warp shuffle allows for fast parallel segmented reduction of cells of varying counts.\\
$\Rightarrow$ Solves warp divergence and shared memory issues mostly, thread groups now correspond to particles. Faster in these instances but NVIDIA only.

\end{frame}

>>>>

\begin{frame}
Sorting can dramatically increase workload performance of subsequent steps for neighboring queries:
\begin{enumerate}
  \item Deep sorted accesses are now \textbf{coalesced}. \label{it:sort_1}
  \item \textbf{Data reuse} due to L2-Cache and/or shared memory. \label{it:sort_2}
\end{enumerate}
\textbf{Double buffer particles} to use last sorted state as input for new sorting to profit from item 1 and 2!
\begin{table}
  \footnotesize
  \begin{tabular}{ | l | r | r | r | r| r |r|}    \hline
    Ordering            &  $\Delta t_c (\mu s)$ & Speedup &VRAM              & SM      & L2     & L2-Hit \\\hline
    Random              &  1,516                & -       &25.0\%            & 3.4\%   & 9.1\%  & 10.8\% \\\hline
    Deep sorted           &    218                & 6.95x   &\textbf{75.3\%}   & 24.4\%  & \textbf{35.0\%} & \textbf{37.8\%} \\
    \hline
 \end{tabular}
\caption{\footnotesize Order dependency of binning of $1024\times1024$ randomly positioned particles in a $128\times128\times128$ grid.}\label{table:bin}
\end{table}
\end{frame}
\subsection{The MPM Specific Transfers}
\begin{frame}
\frametitle{The MPM Specific Transfers}
All MPM operations belong to one of those parallelization schemes:
\begin{itemize}
 \item 1 thread : 1 particle: $\square_p = \square_p \circ \square_p \circ ... \circ \square_p$.
 \item 1 thread : 1 node: $\square_p = \square_p \circ \square_p \circ ... \circ \square_p$.
 \item G2P-transfer: $\square_p = \sum_i \square_i \circ \square_{ip}.$
 \item P2G-transfer: $\square_i = \sum_p \square_p \circ \square_{ip}.$
\end{itemize}
\vfill
MPM-Transfers are executed \textbf{multiple times per physical frame} with varying numbers of variables and mathematical operations.\\\vspace{5}
$\Rightarrow$ Preprocessing steps only need to be done \textbf{once per physical frame}. Sorting already introduced as one of these.
\end{frame}
\begin{frame}
\begin{minipage}{\textwidth}
  G2P-Transfer: $\square_p = \sum_i \square_i \circ \square_{ip}.$\\
  \includesvg[svgpath =../images/,width=\textwidth]{block_transfer_inv_par}
  Pull-Operations read!
\end{minipage}
\end{frame}
\begin{frame}
Similar to filter or stencil operations on the GPU:
\begin{itemize}
  \item 1 node : 1 thread, split grid into blocks corr. to thread groups.
  \item Interpolation function however dependent on particle position.
  \item Needs to be rerun for every particle in the cell.
\end{itemize}
\vfill
Typical setup of transfers:
\begin{enumerate}
  \item Initialize shared memory (pull: with nodes from global memory).
  \item Perform transfers.
  \item Write back to global memory (push: global atomics since writes on halo).
\end{enumerate}
\end{frame}
\begin{frame}
\begin{minipage}{\textwidth}
  P2G-transfer: $\square_i = \sum_p \square_p \circ \square_{ip}.$\\
  \includesvg[svgpath =../images/,width=\textwidth]{block_transfer_par}
  Pull-Operations read, Push-operations write!\\
  $\Rightarrow$ Race conditions.
\end{minipage}
\end{frame}
\begin{frame}
  Typically simulation domain(grid) much bigger than simulation model. Filtering inactive blocks as a preprocess improves performance.
\begin{itemize}
  \item Block is active if any cell counter is active.
  \item Cell counter is active if it has at least one particle.
\end{itemize}
\vfill
\begin{minipage}{0.9\textwidth}
  \footnotesize
  \includesvg[svgpath =../images/,width=\textwidth]{indirect_dispatch}
\end{minipage}
\end{frame}
\begin{frame}
\begin{itemize}
  \item Block and halo are always target of shared memory operations $\Rightarrow$ P2G-pull low occupancy.
  \item \textbf{Batching} multiple particles can increase performance due to hiding synchronization, unroll!
  \item Transfers respect shared memory bank conflicts fully.
  \item \textbf{Warp divergence} for varying cell counts.
\end{itemize}

@@

Simultaneously being worked on. Largely same decision making:\\
\begin{table}
\begin{tabular}{|c|c|c|}
    \hline
	   & \textbf{Me}              & \textbf{\cite{Gao:2018}} \\\hline
  \textbf{Sort}   & Count/Histogram & Count/Histogram \\
	   & for each var.   & sel. variables  \\\hline
  \textbf{Filtering domain}& Filter-op.& Sparse Grid structure \\\hline
  \textbf{Transfers}        & Shared mem. only & Warp-shuffle op.\\\hline
  \end{tabular}
\end{table}
\textbf{Warp shuffle} allows for fast parallel segmented reduction of cells of varying counts.\\
$\Rightarrow$ Solves warp divergence and shared memory issues mostly, thread groups now correspond to particles. Faster for varying counts but \textbf{NVIDIA only}.
\end{frame}
\begin{frame}
  \centering{\textbf{Thank you for your attention!}}
  \vfill
\begin{itemize}
  \item \textbf{Bachelor thesis}: \url{https://github.com/MeyerFabian/snow}
  \item \textbf{Master thesis}: \url{https://github.com/MeyerFabian/msc}
  \item \textbf{Presentation}: \url{https://github.com/MeyerFabian/msc/pres}
      \item \textbf{Code} (right now): \url{https://github.com/mpm-msc/snow}
\end{itemize}
\vfill
\centering
\textbf{Questions?}
\end{frame}
